---
title: "Air Quality Index Exploration"
output: word_document
---
##Exploring Air Quality in Philadelphia

This document aims to explore and visualise the Air Quality Index Days Data file, which is data collected about the air quality in Philadelphia over the past ~15 years.

**Author: Tinashe M. Tapera**  
**Data Source: OpenDataPhilly.org**

###Reading in data
```{r reading in data}
setwd("~/Downloads")
library(foreign)
df = read.csv("Air_Quality_Index_Days.csv")
head(df)
```

The data shows how many "good", "moderate", and "unhealthy" air quality days we have had in Philadelphia since 1990 up until 2014. Let's see how many of each there have been.

###Summarisation
```{r, summarising the data}
colSums(df[,2:4])
summary(df[,2:4])
boxplot(df[,2:4], ylab = "Number of Days", xlab = "Type of day", col = c("green", "yellow", "red"))
```

We can see from this summarisation that overall, moderate has been the most common type of day. Let's visualise the data over time.

```{r visualising, include = FALSE}
library(ggplot2)
```
###Plot
```{r plot}
ggplot(df)+geom_line(aes(x = df$Year, y = df$Good, col = "Good"))+geom_line(aes(x = df$Year, y = df$Moderate, col = "Moderate"))+geom_line(aes(x = df$Year, y = df$Unhealthy, col = "Unhealthy"))+labs(x = "Year", y = "Number of Days", title="Air Quality Index from 1990 - 2014")
```
    
From this plot we can see that the number of unhealthy days has decreased since 1990, while the number of good days has gone up. Moderate days have also gone up, but not by much. Perhaps the next steps would be to:  
1. Investigate the changes in Asthma prevalence over the same time frame;  
2. Investigate where these monitors are;  
3. Understand what Air Quality Index Means.  